---
title: "Factors associated with coverage of deworming in Kenya."
author: "Rasmo Wanyama"
date: "2023-07-21"
output: 
  bookdown::html_document2: default
  bookdown::pdf_document2: default
  bookdown::word_document2: default
toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

# Introduction

We have been given a dataset which contains monthly data for children <5 years, disaggregated at a county level for the period January 2021 to June 2023.

The dataset contains the following variables:

- Period (months from January 2021 to June 2023)
- County (the 47 counties in Kenya)
- Total number of children dewormed (Total Dewormed)
- Number of children <5 years with acute malnutrition (Acute Malnutrition)
- Number of children stunted (0-6 months, 6-23 months, 24-59 months)
- Number of children <5 years with diarrhea (Diarrhea cases)
- Number of children who are underweight (0-6 months, 6-23 months, 24-59 months)

We would like to first explore the data before we formulate a research question.

\newpage

# Data Loading and Exploration

We start by loading important libraries then obtaining the data from a CSV file.
```{r}
## There are two ways of importing packages- you can either use 'pacman' package which automatically installs the packages if they do not exist in the computer

# Importing Packages
library(pacman)
p_load(
  dplyr,
  knitr,
  kableExtr,
  tidyverse,
  scales,
  MASS,
  naniar,
  knitr,
  kableExtra,
  summarytools,
  sf,
  corrplot,
  recipes,
  caret,
)


# The second option of importing packages is using the library() or require() function
library(tidyverse) # this is for importing several packages
library(sf) # this is for importing spatial data (shapefiles)

# Reading the CSV file
data <- read.csv("data/child_health_analysis.csv")

# Overview of the dataset
head(data)
```
```{r}
# View the dataset
View(data)
```

The columns contain county level data on some health indicators for children below 5 years. Specifically,these are:
- Number of children dewormed 
- Number of children with acute malnutrition
- Number of children stunted
- Number of children underweight

```{r}
# Dimensions of the dataset
dim(data)
```
The dataset contains 1410 rows and 11 columns.

```{r}
# Summary statistics
summary(data)
```

To begin our analysis, the first thing we need to do is combine the data on stunting and underweight into one column for children <5 years, and change the period column to a date:

```{r}
# Calculate the sum of 'stunted' cases per row and store it in a new column named 'stunted'
data <- data %>%
  rowwise() %>%
  mutate(stunted = sum(c_across(contains("stunted")), na.rm = TRUE)) %>%

  # Calculate the sum of 'underweight' cases per row
  # and store it in a new column named 'underweight'
  mutate(underweight = sum(c_across(contains("Underweight")), na.rm = TRUE)) %>%

  # Change the date format to one that R understands
  mutate(period = my(period)) %>%

  # Remove the redundant word "County" from the 'county' column
  mutate(county = str_remove(county, "County")) %>%

  # Remove leading and trailing white spaces from the 'county' column
  mutate(county = trimws(county)) %>%

  # Clean the column names using the janitor package
  janitor::clean_names() %>%

  # Select specific columns for analysis
  dplyr::select(period, county, total_dewormed, acute_malnutrition, diarrhoea_cases, stunted, underweight) %>%

  # Remove any row grouping that may exist
  ungroup()

```

This code prepares the data for further analysis by consolidating information on stunting and underweight, formatting dates, and cleaning column names.

Also, sometimes, conflicts arise when you have functions of the same name in different packages. To avoid this conflict I specified the package name before the function, such as janitor::clean_names()


```{r missing values}

# Checking for Missing Values in the Dataset

# Calculate the number of missing values in each column
missing_values_by_column <- colSums(is.na(data))

# Calculate the percentage of missing values in each column (rounded to 2 decimal places)
missing_values_percentage_by_column <- round((missing_values_by_column / nrow(data)) * 100, 2)

# Create a DataFrame to store the results
missing_values_summary <- data.frame(Column = names(missing_values_by_column),
                                     Missing_Values = missing_values_by_column,
                                     Percentage = paste(missing_values_percentage_by_column, "%"))
# # Print the summary
missing_values_summary
```

The "Acute Malnutrition" column has 355 missing values (NA) out of the 1,410 records accounting for 25.18%. This suggests that the data for acute malnutrition is incomplete for some counties or time periods.

The columns "stunted 6-23 months," "stunted 0-<6 months," and "stunted 24-59 months" have 11 (0.78%), 19(1.35%), and 14(0.99%) missing values, respectively. This indicates that the data for stunted children is missing for certain counties or age groups.

The "diarrhoea cases," "Underweight 0-<6 months," "Underweight 6-23 months," and "Underweight 24-59 Months" columns do not have any missing values. This implies that data is available for all counties and age groups in these columns.

```{r}

# Calculate the number of missing values for 'Acute Malnutrition' per county
missing_acute_malnutrition <- data %>%
  group_by(county) %>%
  summarize(Missing_Values = sum(is.na(acute_malnutrition))) %>%
  
  # Filter out counties with zero missing values
  filter(Missing_Values > 0) %>% 
  
  # Sort in descending order
  arrange(desc(Missing_Values)) 

# Add a title and format the table using kable
missing_acute_malnutrition %>%
  kable(caption = "Counties with Missing Values in Acute Malnutrition") %>%
  kable_styling(full_width = FALSE)

```

- Elgeyo Marakwet County, Kericho County, and Lamu County each have the highest number of missing values in the "Acute Malnutrition" column, with 30 missing values each.

- These counties lack data on Acute Malnutrition across the 30-month period from January 2021 to June 2023.
Uasin Gishu County, Nyandarua County, Bomet County, Kisii County, Nyamira County, and Kirinyaga County also have a substantial number of missing values, ranging from 20 to 30.

- Counties like Samburu, Kitui, Trans Nzoia, Kakamega, Muranga, Tharaka Nithi, Baringo, Kilifi, Embu, and Laikipia have relatively fewer missing values, ranging from 1 to 6.


From this missing value observation, the presence of missing values in the dataset could carry various implications depending on the specific county. For instance, in certain counties, these missing values might indicate that no observation was recorded or that the actual count was zero. This suggests the complete absence of data for those particular counties. Therefore, dropping the missing values in the "Acute Malnutrition" column is the best approach for handling missing values since imputing them with measures like mean, median, or mode for this counties has been deemed impractical.

```{r}

# Drop all rows with missing values in the entire dataset
cleaned_data <- na.omit(data)
missing_values <- sum(is.na(cleaned_data))
missing_values

```

All the missing values in the dataset have been successfully handled.

```{r}
# Calculate annual averages for selected columns
annual_averages <- cleaned_data %>%
  
  # Extract the year from the 'period' column
  mutate(year = str_sub(period, 1, 4)) %>%

  # Group the data by year
  group_by(year) %>%

  # Calculate the mean (average) for each column of interest
  summarise(
    mean_dewormed = mean(total_dewormed),  # Updated column name
    mean_diarrhoea = mean(diarrhoea_cases), 
    mean_stunted = mean(stunted),
    mean_underweight = mean(underweight)
  ) %>%

  # Round the mean values to integers
  mutate(across(contains("mean"), round))

# Convert annual_averages to a dataframe
annual_averages_df <- as.data.frame(annual_averages)

# Rename the columns for better clarity
colnames(annual_averages_df) <- c("Year", "Dewormed", "Diarrhea", "Stunted", "Underweight")

# Print the resulting dataframe
annual_averages_df

```

The annual averages for children under 5 years in Kenya between 2021 and 2023 reveal important insights into child health and well-being.
- In 2021, an average of 11,534 children were dewormed annually, while 2,807 cases of diarrhea were reported. Additionally, 496 children were found to be stunted, indicating growth issues, and 1,102 were underweight. 

- In 2022, the number of children dewormed increased to 13,808, with 3,044 cases of diarrhea, 628 stunted children, and 1,427 underweight children. 

- By 2023, the annual averages continued to rise, with 13,898 children dewormed, 3,557 cases of diarrhea, 831 stunted children, and 1,869 underweight children. These statistics underscore the ongoing efforts to address child health challenges in Kenya and highlight areas where further interventions may be needed.

```{r descriptive statistics}

descriptive_stats1 <- combined_data1 %>%
  mutate(year=str_sub(period, 1,4)) %>% # create a column with the period rounded off to a year
  group_by(year)%>%
  dplyr::summarise(mean_dewormed=mean(total_dewormed),
           mean_diarrhoea= mean(diarrhoea_cases),
            mean_stunted=mean(stunted),
            mean_underweight=mean(underweight)) %>% # calculate the mean cases per year
  mutate(across(contains("mean"), round)) # round off the mean values to integers

#Table 1
descriptive_stats1 %>%
  kable(digits = 2, caption = "Annual average number of cases reported and treatment administered among children <5 years in Kenya between 2021 to 2023", col.names = c("Year", "Dewormed", "Diarrhea", "Stunted", "Underweight"))

```
\newpage

# Data Visualization

We now visualise the trend in the average number of cases reported every month from January 2021 to June 2023, and the treatment data (deworming). The deworming data, (brown points), is interpreted using the secondary y axis on the right. We observe an overall increasing trend in the reported cases of diarrhea, stunting and underweight, and also in the number of children dewormed. 


```{r, fig.width=14, fig.height=8, message=F}

descriptive_stats2 <- combined_data1 %>%
  group_by(period)%>%
  dplyr::summarise(avg_dewormed=mean(total_dewormed),
           mean_diarrhoea= mean(diarrhoea_cases),
            mean_stunted=mean(stunted),
            mean_underweight=mean(underweight)) %>% # calculate the mean cases per year
  mutate(across(contains("mean"), round)) %>% # round off the mean values to integers
  pivot_longer(contains("mean"), names_to="cases", values_to="values") %>% # change the data to a long format
  mutate(cases=recode(cases,"mean_diarrhoea"="Diarrhea", 
        
         "mean_stunted"="Stunted",
         "mean_underweight"="Underweight")) %>%
   ungroup() # always good practice to ungroup data


ggplot(descriptive_stats2, # plot the cases first, without the treatment data
       aes(x=period))+geom_smooth(aes(y=values, color=cases))+geom_point(aes(y=values, color=cases))+ # plot for both the line and point graphs
 theme_bw()+ # choose the theme
  geom_smooth(aes(y=avg_dewormed/10), color="#a65628")+ # add the deworming data as a secondary axis
  geom_point(aes(y=avg_dewormed/10), color="#a65628")+ # add the points to showcase the deworming data
  scale_y_continuous(sec.axis=sec_axis(~.*10, name="Number of children <5 years who were dewormed"))+
  theme(axis.title.y.right = element_text(color = "#a65628"))+ # add a secondary axis
  scale_x_date(date_breaks = "3 months", date_labels= "%b-%Y")+ # change the date breaks to three months (x axis), and the labels
  labs(x="Date (month-Year)", y="Number of syndromes or cases reported in children <5 years", color="")+# change the labels for the x and y axis, together with the legend title.
  scale_color_brewer(palette = "Paired") # change the color palette

  
```

## Choropleth map


This section showcases the prevalence of various malnutrition conditions in Kenya through an interactive choropleth map. The map allows users to hover over each county to access detailed information on specific health indicators. For an optimal viewing experience, it is recommended to knit the document in HTML format. 

Figure \@ref(fig:figure1) represents a choropleth map showing the various variables in the data set. First the data was grouped according to counties and then the sum was obtained. Also, the shape file was imported and was arranged according to the alphabetical order of the county column so that the shape file can be arranged as the original data. Then the geometry column was added.


```{r}

descriptive_stats3 <- combined_data1 %>%
  mutate(year=str_sub(period, 1,4)) %>%
  group_by(year, county)%>%
  dplyr::summarise(mean_dewormed=mean(total_dewormed),
           mean_diarrhoea= mean(diarrhoea_cases),
            mean_stunted=mean(stunted),
            mean_underweight=mean(underweight)) %>% # calculate the mean cases per year
  mutate(across(contains("mean"), round)) %>% # round off the mean values to integers
  pivot_longer(contains("mean"), names_to="cases", values_to="values") %>% # change the data to a long format
  mutate(cases=recode(cases,"mean_diarrhoea"="Diarrhea", 
         "mean_dewormed"="Dewormed",
         "mean_stunted"="Stunted",
         "mean_underweight"="Underweight")) %>%
  ungroup() # always good practice to ungroup data



# Importing the shape file
# To ensure your shapefile reads, download all the files in the shapefile data and store them in the same folder. If you only download the .shp file, the shapefile will not open.
county_shapefile <- st_read("shapefiles/county.shp")

# Combine the shapefile dataset with the dataframe containing county data using the unique identifier between the two datasets which is the county name.
county_data <- full_join(county_shapefile, descriptive_stats3, by=c("Name"="county")) 



```


When we plot the data, we are not able to see the county differences due to the unfavourable scale we have due to the high number of children dewormed. To unmask the differences, we use breaks to categorise the data.

```{r,fig.width=14, fig.height=8}

ggplot(county_data, aes(fill=values))+
  geom_sf()+facet_grid(year~cases)+theme_void()+
  scale_fill_gradient2(low="#ffeda0", high="#b10026")+ # add a color gradient 
  ggsn::north(county_shapefile) # add the North symbol in the data






```


There are many methods of choosing breaks, but for this data, we will use quantiles.

```{r}
# Add breaks into the data
library(classInt) # package for choosing breaks
breaks <- classIntervals(county_data$values, n=5, style = "quantile")

breaks


# we now add breaks to the data
county_data <- county_data%>%
  mutate(values_brks=case_when(
    (values<433)~ "< 432",
    (values>432 & values<=1014) ~ "432 - 1014",
    (values>1014 & values<=2254) ~ "1015 - 2254",
    (values>2254 & values<=5877) ~ "2254 - 5877",
    (values>5877) ~ ">5877"
  ))

# relevel the categories
county_data$values_brks <- fct_relevel(county_data$values_brks, "< 432","432 - 1014", "1015 - 2254","2254 - 5877",">5877" )


ggplot(county_data, aes(fill=values_brks))+
  geom_sf()+facet_grid(year~cases)+theme_void()+
  scale_fill_brewer(palette="YlOrRd")+
  labs(fill="Average number of children <5 years")




```

We observe that:
- There has been a relatively high number of children dewormed in the country, except in the Coastal and Western region.
- A high number of stunting and underweight cases is observed in the arid and semi-arid areas.

To further explore these differences, we would then calculate the number of cases observed per 100,000 children to help us have a standard way of comparing the counties. This is mainly calculated using the population data together with the number of observed cases.



## Research question

From our exploratory data analysis, we have observed cases of diarrhea, stunting and underweight, and relatively high number of children dewormed in the different counties and years. We would then want to assess: 
- **what are the factors associated with coverage of deworming in Kenya?** Here, we will use a mixed effects model to answer the question.

But first, we need to include the population data to be able to calculate the coverage of deworming in every county, and the incidence of cases per 100,000 children <5 years.

```{r}
# We will include the 2019 Kenya census data which has an R package
install.packages("remotes")
remotes::install_github("Shelmith-Kariuki/rKenyaCensus")

library(rKenyaCensus)
population <- V3_T2.3 %>%
  mutate(Age=as.numeric(Age)) %>%
  filter(Age<5 & Age>1 & SubCounty%in%"ALL") %>% # calculate the proportion of children dewormed, which targets children above 1 year
  mutate(deworm=sum(Total)) %>%
  ungroup() %>%
  filter(Age<5 & SubCounty%in%"ALL") %>%
  group_by(County) %>%
  mutate(Total=sum(Total)) %>%
  ungroup() %>%
  dplyr::select(County, Total, deworm) %>%
  distinct() %>%
  mutate(County=str_to_title(County)) %>%
  mutate(County=recode(County, "Elgeyo-Marakwet"="Elgeyo Marakwet", "Murang'a"="Muranga", "Taita/ Taveta"="Taita Taveta", "Tharaka-Nithi"="Tharaka Nithi")) #recoding the names that are different from the dataset above which contains the number of cases. To check if there are differences in the unique columns, you may use this code: setdiff(population$County, county_data$Name)


# model dataset
model_data <- combined_data1 %>%
  full_join(population, by=c("county"="County")) %>%
  mutate(total_dewormed=total_dewormed/deworm*100000, # calculate the coverage per 100,000
         diarrhoea_cases=diarrhoea_cases/Total*100000, # calculate the incidence per 100,000
         stunted=stunted/Total*100000, # calculate the incidence per 100,000
         underweight=underweight/Total*100000)%>% # calculate the incidence per 100,000
  mutate_at(vars(total_dewormed, diarrhoea_cases, stunted, underweight), round) # round off
```

Start with visualizing our data- we observe similar trends

```{r, fig.width=16, fig.height=7}

model_EDA <- model_data %>%
  group_by(period)%>%
  dplyr::summarise(avg_dewormed=mean(total_dewormed),
           mean_diarrhoea= mean(diarrhoea_cases),
            mean_stunted=mean(stunted),
            mean_underweight=mean(underweight)) %>% # calculate the mean cases per year
  mutate(across(contains("mean"), round)) %>% # round off the mean values to integers
  pivot_longer(contains("mean"), names_to="cases", values_to="values") %>% # change the data to a long format
  mutate(cases=recode(cases,"mean_diarrhoea"="Diarrhea", 
        
         "mean_stunted"="Stunted",
         "mean_underweight"="Underweight")) %>%
   ungroup() # always good practice to ungroup data


ggplot(model_EDA, # plot the cases first, without the treatment data
       aes(x=period))+geom_smooth(aes(y=values, color=cases))+geom_point(aes(y=values, color=cases))+ # plot for both the line and point graphs
 theme_bw()+ # choose the theme
  geom_smooth(aes(y=avg_dewormed/10), color="#a65628")+ # add the deworming data as a secondary axis
  geom_point(aes(y=avg_dewormed/10), color="#a65628")+ # add the points to showcase the deworming data
  scale_y_continuous(sec.axis=sec_axis(~.*10, name="Number of children <5 years who were dewormed"))+
  theme(axis.title.y.right = element_text(color = "#a65628"))+ # add a secondary axis
  scale_x_date(date_breaks = "3 months", date_labels= "%b-%Y")+ # change the date breaks to three months (x axis), and the labels
  labs(x="Date (month-Year)", y="Number of syndromes or cases reported in children <5 years", color="")+# change the labels for the x and y axis, together with the legend title.
  scale_color_brewer(palette = "Paired") # change the color palette

```

```{r,fig.width=16, fig.height=14}
model_EDA2 <- model_data %>%
  mutate(year=str_sub(period, 1,4)) %>%
  group_by(year, county)%>%
  dplyr::summarise(mean_dewormed=mean(total_dewormed),
           mean_diarrhoea= mean(diarrhoea_cases),
            mean_stunted=mean(stunted),
            mean_underweight=mean(underweight)) %>% # calculate the mean cases per year
  mutate(across(contains("mean"), round)) %>% # round off the mean values to integers
  pivot_longer(contains("mean"), names_to="cases", values_to="values") %>% # change the data to a long format
  mutate(cases=recode(cases,"mean_diarrhoea"="Diarrhea", 
         "mean_dewormed"="Dewormed",
         "mean_stunted"="Stunted",
         "mean_underweight"="Underweight")) %>%
  ungroup() # always good practice to ungroup data



# Importing the shape file
# To ensure your shapefile reads, download all the files in the shapefile data and store them in the same folder. If you only download the .shp file, the shapefile will not open.
county_shapefile <- st_read("shapefiles/county.shp")

# Combine the shapefile dataset with the daraframe containing county data using the uning the unique identifier among the two datasets which is the county name.
county_data1 <- full_join(county_shapefile, model_EDA2, by=c("Name"="county")) 

# Add breaks into the data
library(classInt) # package for choosing breaks
breaks <- classIntervals(county_data1$values, n=5, style = "quantile")

breaks


# we now add breaks to the data
county_data1 <- county_data1%>%
  mutate(values_brks=case_when(
    (values<410)~ "< 410",
    (values>409 & values<=1129) ~ "410 - 1129",
    (values>1129 & values<=2052) ~ "1130 - 2052",
    (values>2052 & values<=4920) ~ "2052 - 4920",
    (values>4920) ~ ">4920"
  ))

# relevel the categories
county_data1$values_brks <- fct_relevel(county_data1$values_brks, "< 432","432 - 1014", "1015 - 2254","2254 - 5877",">5877" )


ggplot(county_data1, aes(fill=values_brks))+
  geom_sf()+facet_grid(year~cases)+theme_void()+
  scale_fill_brewer(palette="YlOrRd")+
  labs(fill="Average number of children <5 years per 100,000")

```

We now go ahead to the statistical model. Our outcome variable is coverage of deworming. 
We will need to check if the data is skewed. There are many ways to do this- in our case we will use a histogram

```{r, message=F, fig.wigth=14, fig.height=8}

ggplot(model_data, aes(total_dewormed))+
  geom_histogram()+theme_bw()

```

We observe that our data is skewed. We will then use a negative binomial model.


We first conduct univariable model analysis.

```{r, message=F}
library(pglm)
model1a<-  pglm(total_dewormed~diarrhoea_cases, data=model_data, index=c("period", "county"), family=negbin,model="within")

cbind(round(exp(model1a$estimate),5), round(exp(confint(model1a)),5))
summary(model1a)
```
```{r, message=F}

model1b<-  pglm(total_dewormed~stunted, data=model_data, index=c("period", "county"), family=negbin,model="within")

cbind(round(exp(model1b$estimate),5), round(exp(confint(model1b)),5))
summary(model1b)
```

```{r, message=F}

model1c<-  pglm(total_dewormed~underweight, data=model_data, index=c("period", "county"), family=negbin,model="within")

cbind(round(exp(model1c$estimate),5), round(exp(confint(model1c)),5))
summary(model1c)
```

We observe that the confidence intervals of the results from the univariable models are between one, show the independent variables have a $P$ value of less than 0.2 hence we will include them all in the multivariable model.

```{r, message=F}
## multivariable model
model1<-  pglm(total_dewormed~diarrhoea_cases+stunted+underweight, data=model_data, index=c("period", "county"), family=negbin,model="within")


model1_results <- cbind(round(exp(model1$estimate),5), round(exp(confint(model1)),5)) 
model1_results<- data.frame(model1_results)
model1_results$names <- rownames(model1_results)
names(model1_results) <- c("IRR", "low_ci", "high_ci", "names")

# remove the results for the intercept
model1_results1 <- model1_results%>%
  filter(names!="(Intercept)") %>%
  mutate(names=str_to_title(names))
ggplot(model1_results1, aes(x=names, y=IRR))+
  geom_errorbar(aes(ymin=low_ci, ymax=high_ci), size=1)+
  coord_flip()+theme_bw()+geom_hline(yintercept = 1, linetype=2, color="red")+
  labs(y="Incidence Rate Ratios", x="", color="")+
  theme(text=element_text(size=16, face="bold"))+scale_color_brewer(palette="Set1")
  

```

From these analysis, we may observe that the coverage of deworming was significantly associated (though weakly) with:
- An increased risk in stunted cases per 100,000
- A decreased risk in diarrhea cases per 100,000

There is no association between deworming and the incidence of underweight in children.

The result of an increased risk in stunting may be explained by the fact that these children may already be stunted by the time they are enrolled into the deworming program (as deworming starts from 2 years) and the effect of the intervention may not be seen in a short term period.
